---
title: "DADA2 Pipeline"
author: "Fateema Bazzi"
date: "2026-08-01"
output: html_document
---

```{r setup, include=FALSE}
#TODO: make figures prettier
knitr::opts_chunk$set(echo = TRUE)
```

## Include Libraries and Set Working Directory

```{r, echo=FALSE}
setwd(this.path::here())
library(dada2)
library(ggplot2)
library(tidyverse)
```

## Data Retrieval

```{r}
path <- "../data/raw_fastq/unzipped" # directory containing the unzipped fastq files.
list.files(path)
fastqs <- sort(list.files(path,  pattern=".fastq", full.names = TRUE))
# Extract sample names
sample_names <- sapply(strsplit(basename(fastqs), "\\."), `[`, 1)
```

## Read Quality Profiles

```{r}
qual_profs<-plotQualityProfile(fastqs)
qual_profs
```

## Filter and Trim

```{r}
# Place filtered files in filtered/ subdirectory
filts <- file.path(path, "filtered", paste0(sample_names, "_filt.fastq.gz"))
names(filts) <- sample_names

#trunclen determined by previous multiqc and quality profile
#TODO : play w maxEE and trunclen
out <- filterAndTrim(fastqs, filts, truncLen=c(250),
              maxN=0, maxEE=2, truncQ=2, rm.phix=TRUE,
              compress=TRUE) 
head(out)
```

## Learn the Error Rates

```{r}
err <- learnErrors(filts)
```
```{r}
#visualize estimated error rates
err_plot <- plotErrors(err, nominalQ=TRUE)
err_plot
```

## Sample Inference
```{r}
dadas <- dada(filts, err=err)
dadas[[1]]
```

## Construct Sequence Table

```{r}
seqtab <- makeSequenceTable(dadas)
dim(seqtab)
```
```{r}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab))) #indicates all sequences are the exact same length, 250
```

## Remove Chimeras
```{r}
seqtab_nochim <- removeBimeraDenovo(seqtab, method="consensus", verbose=TRUE)
dim(seqtab_nochim)
#proportion on nonchimera reads
sum(seqtab_nochim)/sum(seqtab)
```

## Sequencing Depth Distribution
```{r}
#reads per sample
sample_depth <- rowSums(seqtab_nochim)
head(sample_depth)
summary(sample_depth)
#convert to dataframe
depth_df <- data.frame(
  Sample = names(sample_depth),
  Reads = sample_depth
)
#plot
seq_depth_plot <- ggplot(depth_df, aes(x = Sample, y = Reads)) +
  geom_col() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  ) +
  labs(
    x = "Sample",
    y = "Reads after chimera removal",
    title = "Sequencing Depth per Sample"
  )
seq_depth_plot
```


## Track Reads Through the Pipeline

```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadas, getN), rowSums(seqtab_nochim))

colnames(track) <- c("input", "filtered", "denoised", "nonchim")
rownames(track) <- sample_names
head(track)
```

Visualize
```{r}
#convert into long format
track_long <- track |>
  as.data.frame() |>
  tibble::rownames_to_column("Sample") |>
  pivot_longer(
    cols = -Sample,
    names_to = "Step",
    values_to = "Reads"
  )

#calculate avg number of reads retained
track_summary <- track_long |>
  group_by(Step) |>
  summarize(
    MeanReads = mean(Reads),
    SD = sd(Reads)
  )

#order the steps correctly
track_summary$Step <- factor(
  track_summary$Step,
  levels = c("input", "filtered", "denoised", "nonchim")
)

#plot
retention_bar <- ggplot(track_summary, aes(x = Step, y = MeanReads)) +
  geom_col() +
  geom_errorbar(
    aes(ymin = MeanReads - SD, ymax = MeanReads + SD),
    width = 0.2
  ) +
  theme_minimal() +
  labs(
    x = "Pipeline Step",
    y = "Mean Reads",
    title = "Read Retention Through DADA2 Pipeline"
  )

#add percentages
track_summary <- track_summary %>%
  mutate(PercentRetained = MeanReads / MeanReads[Step == "input"] * 100)
#plot
retention_per_bar <- ggplot(track_summary, aes(x = Step, y = PercentRetained)) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = (MeanReads - SD) / MeanReads[Step == "input"] * 100,
      ymax = (MeanReads + SD) / MeanReads[Step == "input"] * 100
    ),
    width = 0.2
  ) +
  theme_minimal() +
  labs(
    x = "Pipeline Step",
    y = "Reads Retained (%)",
    title = "DADA2 Read Retention"
  )
retention_bar
retention_per_bar
```


## Assign Taxonomy
I use the RDP reference database, to match what was used by the original authors.
```{r}
taxa <- assignTaxonomy(seqtab_nochim, "../data/raw_fastq/unzipped/tax/rdp_19_toGenus_trainset.fa.gz") #genus
```
Inspect taxonomic assignments
```{r}
taxa_print <- taxa # Removing sequence rownames for display only
rownames(taxa_print) <- NULL
head(taxa_print)
```
visualize
```{r}
tax_summary <- data.frame(
  Rank = colnames(taxa),
  Assigned = colSums(!is.na(taxa)),
  Total = nrow(taxa)
)

tax_summary$Percent <- tax_summary$Assigned / tax_summary$Total * 100

#order the steps correctly
tax_summary$Rank <- factor(
  tax_summary$Rank,
  levels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")
)

taxa_assign_plot <- ggplot(tax_summary, aes(x = Rank, y = Percent)) +
  geom_col() +
  theme_minimal() +
  labs(
    x = "Taxonomic Rank",
    y = "ASVs Assigned (%)",
    title = "Taxonomic Assignment Summary"
  )
taxa_assign_plot
```


## Save Results

```{r}
#asv table
saveRDS(seqtab_nochim, "../results/dada2/seqtab_nochim.rds")
write.csv(seqtab_nochim, "../results/dada2/asv_table.csv")

#taxonomy table
saveRDS(taxa, "../results/taxonomy/taxonomy.rds")
write.csv(taxa, "../results/taxonomy/taxonomy.csv")

#read tracker
saveRDS(track, "../results/dada2/track_reads.rds")
write.csv(track, "../results/dada2/track_reads.csv")

#visuals
ggsave("../results/dada2/quality_profiles.png", plot= qual_profs, width = 12, height = 10, units = "in",)
ggsave("../results/dada2/retention_bar.png", plot= retention_bar)
ggsave("../results/dada2/retention_percent_bar.png", plot= retention_per_bar)
ggsave("../results/dada2/sequencing_depth_distibution.png", plot= seq_depth_plot)
ggsave("../results/taxonomy/taxa_assignment_percent.png", plot= taxa_assign_plot)
```

