---
title: "Differential Abundance"
author: "Fateema Bazzi"
date: "2026-08-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load in Libraries and Set Working Directory

```{r, include=FALSE}
setwd(this.path::here())

library(tidyverse)
library(phyloseq)
library(ggplot2)
library(ANCOMBC)
library(microbiome)
library(RColorBrewer)
library(ggrepel)
```

## Load in Data

```{r}
physeq <- readRDS("../results/phylogeny/phyloseq.rds")
```

## Create Genus-Level Object

```{r}
physeq_genus <- tax_glom(
    physeq,
    taxrank = "Genus"
)
#check how many taxa remain
ntaxa(physeq_genus) #down from 2173
#handle unknowns
taxa_names(physeq_genus)[is.na(taxa_names(physeq_genus))] <- "Unclassified"
```

## ANCOM-BC

```{r}
#set control group
sample_data(physeq_genus)$Group <- relevel(
    sample_data(physeq_genus)$Group,
    ref="EU"
)

#preserve taxon names
taxa_names(physeq_genus) <- paste0(
    "Genus_",
    seq_len(ntaxa(physeq_genus))
)
tax_map <- data.frame(
    Taxon_ID = taxa_names(physeq_genus),
    as.data.frame(tax_table(physeq_genus))
)

da_results <- ancombc(
    data = physeq_genus,
    formula = "Group",
    p_adj_method = "BH" #multiple testing correction
)
```
Extract Log FC and Q-Vals
```{r}
lfc <- da_results$res$lfc
qvals <- da_results$res$q_val
```

Filter out results with adjusted p-value>0.05 and |lfc|<1
```{r}
sig_taxa <- rownames(
    da_results$res$q_val[
        da_results$res$q_val[, "GroupBF"] < 0.05 &
        abs(da_results$res$lfc[, "GroupBF"]) > 1,
    ]
)
```

## Differential Abundance Table
```{r}
da_table <- data.frame(
    Taxon_ID = rownames(da_results$feature_table),
    logFC = da_results$res$lfc[, "GroupBF"],
    p_value = da_results$res$p_val[, "GroupBF"],
    q_value = da_results$res$q_val[, "GroupBF"]
)

#add significance indication
da_table$Significant <- 
    da_table$q_value < 0.05 &
    abs(da_table$logFC) > 1

#add direction of differential abundance
da_table$Direction <- ifelse(
    da_table$logFC > 0,
    "BF enriched",
    "EU enriched"
)

# add taxonomy info
tax_df <- as.data.frame(tax_table(physeq_genus))
tax_df$Taxon_ID <- rownames(tax_df)

da_table <- merge(
    da_table,
    tax_df,
    by = "Taxon_ID",
    all.x = TRUE
)

#add taxon column
da_table$Taxon <- da_table$Genus

#reorder
da_table <- da_table[, c(
    "Taxon",
    "Phylum",
    "Class",
    "Order",
    "Family",
    "Genus",
    "logFC",
    "p_value",
    "q_value",
    "Significant",
    "Direction",
    "Taxon_ID"
)]

head(da_table)
```

## Visualize
Volcano Plot
```{r}
#TODO: Refine
myvolcanoplot <- ggplot(data = da_table, aes(x = logFC, y = -log10(q_value), col = Significant, shape = Direction, label=Taxon)) +
  geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point()+
  geom_text_repel(max.overlaps = Inf)
myvolcanoplot
```

Heat Map
```{r}

```

Bar Plot
```{r}

```

## Save Results
```{r}
#differential abundance table
saveRDS(da_table, "../results/differential_abundance/diff_abund_tbl.rds")
write.csv(da_table, "../results/differential_abundance/diff_abund_tbl.csv")
#visuals
```